Calendar-based graphics for visualising people's daily schedules

### 2017-07-04 main.Rmd
added more on reference lines and labels

---
title: "Calendar-based graphics for visualising people's daily schedules"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Di
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
fontsize: 11pt
papersize: a4
fontfamily: mathpazo
blind: false
cover: true
wpnumber: no/yr
jelcodes: C10,C14,C22
bibliography: references.bib
biblio-style: authoryear-comp
abstract: "A brief summary of our ideas"
keywords: "blah, blah"
subparagraph: true
toc: false
output: 
  bookdown::pdf_document2:
    template: monashwp.tex
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    number_sections: yes
    citation_package: biblatex
---

```{r initial, echo = FALSE, cache = FALSE, results = 'hide'}
library(knitr)
opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE,
  fig.path = 'figure/', cache.path = 'cache/', fig.align = 'center', 
  fig.show = 'hold', cache = TRUE, external = TRUE, dev = "svglite"
)
```

@@

Calendar-based visualisation proves useful in unfolding human-related activities in temporal context. For example, @van_wijk_cluster_1999 developed a calendar view of heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It effectively contrasts weekdays and weekends, highlights public holiday, and presents other known seasons like school vacations, all of which have influence over the turn-outs in the office. However, their technique is too constrained to colour-encoding graphics. We shall extend this calendar-based arrangement using linear algebra tools from a single type of heatmap to a broader range of glyphs, from univariate to multivariate cases, from one type of calendar to three types. The proposed algorithm has been implemented in the `sugrrants` package [@sugrrants] using R [@r].

This work is partly motivated by studying hourly foot traffic in the city of Melbourne [@ped]. There have been 43 sensors installed across downtown Melbourne since 2009 in order to capture the pulse of city life. (FIGURE ?: weekday at multiple sensors) Figure (?) lends itself to addressing some of the issues: (1) to simultaneously exhibit multiple seasons including time of day, day of week, and day of year, (2) to visualise long historical time series data in an information-dense way at the overview level, (3) to compare and contrast multiple time series on a common scale, (4) to quickly look up the date when there are unexpected events happening without interactivity.

@@

(NEED A BRIEF INTRO HERE)

## Grids setup

@@

One of the advantages using glyphs over heatmaps is evident in easily enabling scales on different temporal scales. All of the line glyphs shown above are drawn on the global scale so that it makes the magnitudes comparable over all the period. On the other hand, some sub-series of small magnitudes yet worth-noting shapes possibly become invisible whilst embedded in the overall picture. The individual scale or something else, as a result, helps to complement the calendar-based visualisation in general.

@@

The major reference lines are placed on the far left and the bottom for every month panel: for each $m$, the vertical separation is $\min{(x)}$; for each $n$, the horizontal separation is $\min{(y)}$. The month labels located on the top is obtaining $\max{(y)}$ for each $n$. The minor reference lines for every daily grids are placed on the starting positions: for each $i$, the vertical line is $\min{(x)}$; for each $j$, the horizontal line is $\min{(y)}$.

(INTERACTIVITY COULD BE DISCUSSED HERE) We shall enable interactivity in the calendar-based graphics for time series data. It will allow users to transform different temporal components and switch displays between overlaying and faceting through key strokes or mouse clicks.

# Grammar of graphics

# Conclusions

@@

>>>>

---
title: "Calendar-based graphics for visualising people's daily schedules"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Di
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
fontsize: 11pt
papersize: a4
fontfamily: mathpazo
blind: false
cover: true
wpnumber: no/yr
jelcodes: C10,C14,C22
bibliography: ["references.bib", "packages.bib"]
biblio-style: authoryear-comp
abstract: "A brief summary of our ideas"
keywords: "blah, blah"
subparagraph: true
toc: false
output: 
  bookdown::pdf_document2:
    template: monashwp.tex
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    number_sections: yes
    citation_package: biblatex
---

```{r initial, echo = FALSE, cache = FALSE, include = FALSE}
library(knitr)
opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE,
  fig.path = 'figure/', cache.path = 'cache/', fig.align = 'center', 
  fig.show = 'hold', cache = TRUE, external = TRUE, dev = "svglite"
)
read_chunk('src/main.R')
```

```{r load}
```

@@

Calendar-based visualisation proves useful in unfolding human-related activities in temporal context. For example, @van_wijk_cluster_1999 developed a calendar view of heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It effectively contrasts weekdays and weekends, highlights public holiday, and presents other known seasons like school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in a couple of R packages: `ggTimeSeries` [@R-ggTimeSeries] and `ggcal` [@R-ggcal]. However, their technique is too constrained to colour-encoding graphics. We shall extend this calendar-based arrangement using linear algebra tools from a single type of heatmap to a broader range of glyphs, from univariate to multivariate cases, from one calendar format to three formats. The proposed algorithm has been developed in the `sugrrants` package [@R-sugrrants] using R [@R-base].

This work was originally motivated by studying hourly foot traffic in the city of Melbourne [@ped]. There have been 43 sensors installed across downtown Melbourne since 2009 in order to capture the pulse of city life. (FIGURE ?: weekday at multiple sensors) Figure (?) lends itself to addressing some of the issues: (1) to simultaneously exhibit multiple seasons including time of day, day of week, and day of year, (2) to visualise long historical time series data in an information-dense way at the overview level, (3) to compare and contrast multiple time series on a common scale, (4) to quickly look up the date when there are unexpected events happening without interactivity.

@@

In the section, the algorithms to construct calendar grids, different scales, and reference lines and labels are discussed in detail.

## Grid construction

@@

One of the advantages using glyphs over heatmaps is evident in easily enabling scales on different temporal scales. All of the line glyphs shown above are drawn on the global scale so that it makes the magnitudes comparable over all the period. On the other hand, some sub-series of small magnitudes yet worth-noting shapes possibly become invisible whilst embedded in the overall picture. The individual scale or something else, as a result, complements the calendar-based visualisation in general.

There are other three types of scales in addition to the global scale: individual scales regardless of days, conditional on weekdays, and conditional on days of a month.

@@

It can be noted that reference lines separating each cell and panel and labels indicating weekday and month are also constructed in order to make calendar-based graphics more accessible and informative.

Regarding the monthly calendar format, the major reference lines separate every month panel and the minor ones separate every day cell, represented by the think and thin lines respectively. The major reference lines are placed on the far left and right as well as the bottom and top of every month panel: for each $m$, the vertical lines are determined by $\min{(x)}$ and $\max{(x)}$; for each $n$, the horizontal lines are computed by $\min{(y)}$ and $\max{(y)}$. The minor reference lines are placed on the initial positions: for each $i$, the vertical line is $\min{(x)}$; for each $j$, the horizontal line is $\min{(y)}$.

The abbreviated month labels located on the top left are obtained through $(\min{(x)}, \max{(y)})$ for every $m$ and $n$. The weekday labels are uniformly placed on the bottom of the whole canvas, that is $\min{(y)}$.

# The `frame_calendar` function

```{r args, eval = FALSE, echo = TRUE}
frame_calendar(data, x, y, date, calendar = "monthly", dir = "h",
  sunday = FALSE, nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed"
)
```

# Discussion

@@

We shall enable interactivity in the calendar-based graphics for time series data. It will allow users to transform different temporal components and switch displays between overlaying and faceting through key strokes or mouse clicks.

```{r write-bib}
write_bib(c( .packages()), 'packages.bib')
```
